---
title: "class05"
author: "Sam Fisher A18131929"
format: pdf
toc: true
---

## Background

There are lots of ways to make plots in R. These include so-called "Base R" (like the 'plot()') and add on packages like **ggplot2**

Let's make the same plot with these two graphics systems. We can use the inbuilt 'cars' data set:

```{r}
head(cars)
```

with "base R" we simply:

```{r}
plot(cars)
```

Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`

**N.I.K.** We never run on `install.packages() in a code chunk otherwise we will re-install needlessly every time we open the document

Every time we want to use an add we must load it up using `library()`
```{r}
library(ggplot2)
ggplot(cars)
```


Every ggplot needs at least three things:

1. The **Data** i.e. stuff to plot as data frame
2. the **aes** are aesthetics that map the data to the plot
3. the **geom** or geometry i.e. the plot type such as scatter, linear, etc


```{r}
ggplot(cars) +
  aes(x=speed, y=dist) + geom_point() + geom_line()
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + geom_point() + geom_smooth(method="lm", se=FALSE) + labs(x="speed(mph)", y="Distance(ft)", title="Stopping Distance of Old Cars") + theme_bw()
```





##Gene Expression Plot

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2) + 
  geom_point(col="blue", alpha=0.2)
```

Let's color by `State` up, down or no change

```{r}
table(genes$State)
```

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point() +
  scale_color_manual(values=c("red",
                              "gray",
                              "blue") ) + 
  labs(x="Control (no drug)",
     y= "Drug",
     title= "Expression changes with GLP-1 Drug") 

```


## Going further with gapminder

Here we explore the famous `gapminder` dataset with some custom plots. 


```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many rows does this dataset have?

```{r}
nrow(gapminder) 
```

>Q. How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```


Version 1 plot gdpPercap vs LifeExp for all rows

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point() 
```

I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point() +
  facet_wrap(~continent)
```


## First look at the dplyr package

Another add-on package with a function called `filter()` that we want to use

```{r}
library(dplyr)
```

```{r}
filter(gapminder, year == 2007, country == "Ireland")
```

```{r}
input <- filter(gapminder, year == 2007 | year == 1997)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, color = continent) +
  geom_point() +
  filter(gapminder, year ==2007 | year == 1997) + 
  facet_wrap(~year)
```





