---
title: "Class 18: Pertusssis Mini Project"
author: "Sam Fisher"
format: pdf
---

## Background
Pertussis (aka whooping cough) is a common lung infection caused by the bacteria B. Pertussis.

This can infect people of any age, but it is most lethal to infants due to their smaller airways. 
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html
The CDC tracks the number of reported cases in the US:
We can "scrape" this data with the **datapasta** package

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L, 2024L, 2025L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063, 22538, 21996)
       )
```

> Q. Make a plot of `year` vs `cases`

```{r}
library(ggplot2)
ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() + geom_line()
```

> Q. Add some major milestones including the first wP vaccine roll-out (1946), the switch to the newer aP vaccine (1996), the COVID virus (2020)

```{r}
ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "green")
```

There were high case numbers in the pre 1940s followed by a sharp decline after the introduction of the whole-cell pertussis vaccine in 1946. The spike in the mid 2000s is likely associated with anti-vaccine movements in the media that have caused people to believe vaccines are bad for their health. 

There were high case numbers pre 1940s followed by a sharp decline after the introduction of the whole-cell pertussis vaccine in 1946. Case numbers remained fairly low and constant until the mid 2000s when the ap vaccine was introduced. Countries such as Japan who switched two years before the U.S. show a similar pattern but two years earlier. Other countries that switched later also show the same pattern but later. It appears that the ap vaccine wears off faster. 


**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## CMI-PB Data
[Computational Models of Immunity - Pertussis Boost](https://www.cmi-pb.org/) project aims to provide the scientific community with this very information. 

They make their data available vis JSON formati returning API. We can read this in R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> Q. How mant wP and aP individuals are in this table?

```{r}
table(subject$infancy_vac)
```

> Q. What is the biological sex breakdown

```{r}
table(subject$biological_sex)
```

> Q. In terms of race and gender is this dataset representative of the US population

```{r}
table(subject$race, subject$biological_sex)
```

Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To analyze this data we need to first "join" (merge/link) the different tables so we ahve all ther data in one place not spread accross different tables. 

We can use the `*_join()` family of functions from **dplyr** to do this 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. What antibody isotypes are measured for these patients? 

```{r}
table(abdata$isotype)
```


> Q. What antigens are reported? 

```{r}
table(abdata$antigen)
```

Let's focus on the IgG isotype and make a plot of MFI_normalized for all antigens. 

```{r}
igg <- abdata |>
  filter(isotype == "IgG")

head(igg)
```


```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) + geom_boxplot()

```

> Q. Is there a difference for aP vs wP individuals with these values? 

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col=infancy_vac) + geom_boxplot() 

```

> Q. Is there a temrol response - i.e. do values increase or decrease over time?

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col=infancy_vac) + geom_boxplot() + facet_wrap(~visit)

```

## Focus on "PT" Pertusis Toxin antigen 

```{r}
pt.igg.21 <- igg |> filter(antigen == "PT",
              dataset == "2021_dataset")
```


```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col=infancy_vac,
      group = subject_id) +
  geom_point() + geom_line() 
  geom_vline(xintercept = 14, lty = "dashed")
```

```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col = infancy_vac,
      group = subject_id) +
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.3) +
  geom_smooth(aes(group = infancy_vac), se = FALSE, linewidth = 2) +
  geom_vline(xintercept = 14, lty = "dashed") +
  labs(
    title = "IgG Response to Pertussis Booster Over Time",
    x = "Days Relative to Booster",
    y = "Normalized MFI",
    color = "Infancy Vaccine"
  ) +
  theme_classic()
```




















