---
title: "Class11: AlphaFold"
author: "Samuel Fisher (A18131929)"
format: pdf
---

## Background

In this hands-on session we will utilize AlphaFold to predict protein structure from sequence (Jumper et al. 2021).
Without the aid of such approaches, it can take years of expensive laboratory work to determine the structure of just one protein. With AlphaFold we can now accurately compute a typical protein structure in as little as ten minutes.
The PDB database(the main repositry of experimental structures) only has approximately 250 thousand structures (we saw this in the last lab). The main protein sequence database has over **200 million** sequences! Only 0.125% of known sequences have a known structure - this is called the "structure knowledge gap". 

```{r}
(250000 / 200000000) * 100
```
Structures are much harder to determine than sequences
They are expensive (on average ~$1 million each)
They take on average 3-5 years to solve!

# EBI AlphaFold Database
The EBI has a database of pre-computed AlphaFold (AF) models called AFDB. This is growing all the time and can be useful to check before running AF ourselves. 

## Running AlphaFold

We can download and run locally (on our own computers) but we need a GPU. Or we can use "cloud" computing to run this on somene else's computer

We will use ColabFold < https://github.com/sokrypton/ColabFold

We previously found there was no AFDB entry for our HIV sequence. 

```
>HIV-Pr-Dimer
PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF:PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF
```

Here we will use AlphaFold2_mmseqs2

I was unable to use Mol* to generate the image frm section 7. I was able to view the proteins, but unable to select the individual proteins in order to superimpose them. 


## Section 8

# Section 8: point to YOUR results folder (must contain the 5 .pdb files)
```{r}
results_dir <- "hivpr_23119/"
```

```{r}
## list all PDB models in current project folder
pdb_files <- list.files(pattern="\\.pdb$", full.names=TRUE)

basename(pdb_files)
```

```{r}
library(bio3d)

# Read all PDB models and superpose/fit coordinates
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```


```{r}
pdbs
```


```{r}
rd <- rmsd(pdbs, fit=TRUE)
range(rd)
```

```{r}
library(pheatmap)
# compute RMSD matrix
rd <- rmsd(pdbs, fit = TRUE)
# auto-size labels to match matrix
n <- ncol(rd)
labs <- paste0("m", seq_len(n))
colnames(rd) <- labs
rownames(rd) <- labs
pheatmap(rd)
```

```{r}
library(bio3d)
# Read a reference PDB for secondary structure annotation
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)
```

```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error

```{r}
library(jsonlite)

# list all PAE JSON files (should be 5)
pae_files <- list.files(pattern=".*model.*\\.json$", full.names=TRUE)

basename(pae_files)
length(pae_files)
```

```{r}
pae1 <- read_json(pae_files[1], simplifyVector = TRUE)
pae5 <- read_json(pae_files[5], simplifyVector = TRUE)

attributes(pae1)
```

```{r}
head(pae1$plddt)
```

```{r}
pae1$max_pae
pae5$max_pae
```

```{r}
library(bio3d)

```

```{r}
plot.dmat(pae1$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")

plot.dmat(pae5$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))

plot.dmat(pae1$pae,
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))

```

## Residue conservation from alignment file

```{r}
aln_file <- list.files(pattern="\\.a3m$", full.names=TRUE)
aln_file
```

```{r}
library(bio3d)

aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)
```

```{r}
pdb <- read.pdb(pdb_files[1])
sim <- conserv(aln)

plotb3(sim[1:99],
       sse = trim.pdb(pdb, chain = "A"),
       ylab = "Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])

occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)

write.pdb(m1.pdb, o = occ, file = "m1_conserv.pdb")
```










































