---
title: "Class17 PCA"
author: "Sam Fisher (A18131929)"
format: pdf
---

## Downstream Analysis

```{r}
library(tximport)

# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

We now have our estimated transcript counts for each sample in R. We can see how many transcripts we have for each sample:

```{r}
colSums(txi.kallisto$counts)
```

And how many transcripts are detected in at least one sample:

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filtering out annotated transcripts with no reads:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

And filtering out the ones without any change over the samples

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## PCA Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

# PC1 vs PC2 plot

```{r}
library(ggplot2)
library(ggrepel)
```

```{r}
pc.df <- as.data.frame(pca$x)
pc.df$sample <- rownames(pc.df)
pc.df$group <- c("control", "control", "treatment", "treatment")

ggplot(pc.df, aes(x = PC1, y = PC2, color = group, label = sample)) +
  geom_point() +
  geom_text_repel() +
  theme_minimal()
```

# PC1 vs PC3

```{r}
ggplot(pc.df, aes(x = PC1, y = PC3, color = group, label = sample)) +
  geom_point() +
  geom_text_repel() +
  theme_minimal()
```

# PC2 vs PC3

```{r}
ggplot(pc.df, aes(x = PC2, y = PC3, color = group, label = sample)) +
  geom_point() +
  geom_text_repel() +
  theme_minimal()
```


## Differential-expression Analysis

```{r}
library(DESeq2)
```

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```






















